---
title: "eDNA Analysis"
author: "Nathan Reed (24110024)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Explore the dataset to understand what level of coverage we have for our data sources

```{r}
library(tidyverse)
library(ggplot2)
library(grid)
library(patchwork)
library(VennDiagram)

data2 <- read_csv("../Data/data_enriched.csv")
bayes <- read.csv("../Data/ALL_ROWS_BAYESIAN_DATA.csv")

# If any -Inf values slipped through, set to 0
data2 <- data2 %>%
  mutate(across(c(DNA_score, obs_score), ~ ifelse(is.infinite(.x) & .x < 0, 0, .x)))

bayes <- bayes %>%
  mutate(across(c(DNA_score, obs_score), ~ ifelse(is.infinite(.x) & .x < 0, 0, .x)))

# Clean blunt_score (force numeric, replace NA and non-numeric with 0)
bayes <- bayes %>%
  mutate(blunt_score = suppressWarnings(as.numeric(blunt_score)),
         blunt_score = replace_na(blunt_score, 0))

data2 <- data2 %>%
  mutate(blunt_score = suppressWarnings(as.numeric(blunt_score)),
         blunt_score = replace_na(blunt_score, 0))


str(bayes)
```


## Explore the scores to understand how the underlying data is influencing things

Using the current formula, we have peaks for the blunt_score at 1 and 0, which is good (I think). It means for a large proporation of the samples we're strongly confident that it's right or wrong.

I was hoping in the second plot that there would be a stronger relationship between our blunt_score and the X.ID but there isn't

```{r cars}
hist(data2$blunt_score, breaks=20)
hist(bayes$Combined.Probability, breaks=20)
plot(data2$X.ID,data2$blunt_score)
```

```{r}
hist(data2$geo_evidence_score, breaks=5)
```

```{r}
# The proportion of times each data source is used to determine location validity

# Max probability from the 3 different observation scores # <20k=1, <50k=0.99, <100k=0.90 
nz <- function(x) ifelse(is.na(x), FALSE, x) 
nearest_GBIF <- pmax( 
  ifelse(nz(data2$GBIF_within_20km), 1.00, 0), 
  ifelse(nz(data2$GBIF_within_50km), 0.99, 0), 
  ifelse(nz(data2$GBIF_within_100km), 0.90, 0) ) 

nearest_OBIS <- pmax( 
  ifelse(nz(data2$obis_within_20km), 1.00, 0), 
  ifelse(nz(data2$obis_within_50km), 0.99, 0), 
  ifelse(nz(data2$obis_within_100km), 0.90, 0) ) 

am_score <- ifelse(is.na(data2$am_prob), 0, data2$am_prob)

# Build matrix in the order: Aquamaps (am_score, GBIS, OBIS)
wmat <- cbind(am_score, nearest_GBIF, nearest_OBIS)
wmat[!is.finite(wmat)] <- -Inf
labels <- c("Aquamaps", "GBIF", "OBIS")

# Row-wise max and tie logic
maxv <- apply(wmat, 1, max)
tol  <- 1e-9
is_at_max <- abs(wmat - maxv) <= tol
n_at_max  <- rowSums(is_at_max & is.finite(wmat))

has_positive <- is.finite(maxv) & (maxv > 0)  # at least one positive signal

# Classes
is_winner   <- has_positive & (n_at_max == 1)
is_two_src  <- has_positive & (n_at_max == 2)
is_three_src<- has_positive & (n_at_max == 3)
is_none     <- !has_positive

# Winner labels for single-max rows
idx <- max.col(wmat, ties.method = "last")   # ties will be excluded below
winner <- factor(labels[idx], levels = labels)
winner[!is_winner] <- NA

# Counts
tab_winners <- table(winner, useNA = "no")
two_count   <- sum(is_two_src)
three_count <- sum(is_three_src)
none_count  <- sum(is_none)

# Compose counts: 3 winners + 2 sources + 3 sources + None
counts <- c(Aquamaps = 0L, GBIF = 0L, OBIS = 0L, `2 sources` = two_count, `3 sources` = three_count, None = none_count)
counts[names(tab_winners)] <- as.integer(tab_winners)

# Plot
barplot(counts, main = "Use of data sources for validation sample location", ylab = "Count", xlab = "Data sources")

```

## Comparison of Blunct Score Calculation and Bayesian Posterior Mean

```{r}
# Choose a common binwidth so the two histograms are comparable
bin_w <- 0.02  # adjust to taste

pBlunt <- ggplot(bayes, aes(x = blunt_score)) +
  geom_histogram(binwidth = bin_w, linewidth = 0.2) +
  labs(title = "Blunt score", x = "blunt_score", y = "Count") +
  theme_minimal(base_size = 12)

pBayes <- ggplot(bayes, aes(x = Combined.Probability)) +
  geom_histogram(binwidth = bin_w, linewidth = 0.2) +
  labs(title = "Posterior mean", x = "Combined.Probability", y = "Count") +
  theme_minimal(base_size = 12)

pBlunt + pBayes


```

The Bayesian approach always calculates a mean, whereas we have many missing values for the Blunt approach (approx 17%)

```{r}
avail <- tibble(total = nrow(bayes), blunt_p0 = sum(bayes$blunt_score==0), bayes_p0 = sum(bayes$Combined.Probability==0))
avail

# Compare histograms with 0 values removed

# Make a paired subset (fair comparison)
paired <- bayes %>% select(blunt_score, Combined.Probability) %>% filter(blunt_score>0, Combined.Probability>0)

# Extract summary stats
paired_summary <- paired %>%
  summarise(
    blunt_mean = mean(blunt_score),
    blunt_median = median(blunt_score),
    bayes_mean = mean(Combined.Probability),
    bayes_median = median(Combined.Probability),
    corr = cor(blunt_score, Combined.Probability)
  )

paired_summary

# Side-by-side histograms (paired subset only)
bin_w <- 0.02  # tweak if needed

p_blunt <- ggplot(paired, aes(x = blunt_score)) +
  geom_histogram(binwidth = bin_w, linewidth = 0.2) +
  labs(title = "Blunt score (paired)", x = "blunt_score", y = "Count") +
  theme_minimal(base_size = 12)

p_bayes <- ggplot(paired, aes(x = Combined.Probability)) +
  geom_histogram(binwidth = bin_w, linewidth = 0.2) +
  labs(title = "Posterior mean (paired)", x = "Combined.Probability", y = "Count") +
  theme_minimal(base_size = 12)

p_blunt + p_bayes

```
## Proportin of scores with > 0.9 probability

```{r}
venn_cutoff = 0.8

bayes %>%
  summarise(
    blunt_count = sum(blunt_score > venn_cutoff, na.rm = TRUE),
    blunt_pct = mean(blunt_score > venn_cutoff, na.rm = TRUE),
    bayes_count = sum(Combined.Probability > venn_cutoff, na.rm = TRUE),
        bayes_pct = mean(Combined.Probability > venn_cutoff, na.rm = TRUE)
  )

cmp <- bayes %>% mutate(blunt90 = blunt_score > venn_cutoff, bayes90 = Combined.Probability > venn_cutoff)

# 2×2 contingency table of selections
overlap_tbl <- cmp %>% count(blunt90, bayes90)
overlap_tbl

both        <- sum(cmp$blunt90 & cmp$bayes90, na.rm = TRUE)
blunt_only  <- sum(cmp$blunt90 & !cmp$bayes90, na.rm = TRUE)
bayes_only  <- sum(!cmp$blunt90 & cmp$bayes90, na.rm = TRUE)
neither     <- sum(!cmp$blunt90 & !cmp$bayes90, na.rm = TRUE)

# If you want simple overlap scores (treat Bayes as "reference"):
precision <- ifelse((both + blunt_only) > 0,  both / (both + blunt_only), NA_real_)
recall    <- ifelse((both + bayes_only) > 0, both / (both + bayes_only), NA_real_)
f1        <- ifelse(is.finite(precision + recall) & (precision + recall) > 0,
                    2 * precision * recall / (precision + recall), NA_real_)
jaccard   <- ifelse((both + blunt_only + bayes_only) > 0,
                    both / (both + blunt_only + bayes_only), NA_real_)

tibble(both, blunt_only, bayes_only, neither, precision, recall, f1, jaccard)



# Venn diagram of >0.9 overlap
blunt_only <- overlap_tbl %>% filter(blunt90, !bayes90) %>% pull(n)
bayes_only <- overlap_tbl %>% filter(!blunt90, bayes90) %>% pull(n)
both       <- overlap_tbl %>% filter(blunt90,  bayes90) %>% pull(n)

grid.newpage()
draw.pairwise.venn(
  area1 = blunt_only + both,
  area2 = bayes_only + both,
  cross.area = both,
  category = c("Blunt > 0.5", "Bayes > 0.5")
)

```
## Comparing extreme differences in the predictions


```{r}
# Thresholds for comparison
thr_low  <- 0.1
thr_high <- 0.9

# keep only rows where both values are present
bx <- bayes %>%
  filter(!is.na(blunt_score), !is.na(Combined.Probability)) %>%
  mutate(diff = Combined.Probability - blunt_score)

# Case A: blunt very low, Bayes very high
low_blunt_high_bayes <- bx %>%
  filter(blunt_score <= thr_low, Combined.Probability >= thr_high) %>%
  arrange(desc(diff))

# Case B: blunt very high, Bayes very low
high_blunt_low_bayes <- bx %>%
  filter(blunt_score >= thr_high, Combined.Probability <= thr_low) %>%
  arrange(diff)

# Quick counts so you know what you have
tibble(
  low_blunt_high_bayes = nrow(low_blunt_high_bayes),
  high_blunt_low_bayes = nrow(high_blunt_low_bayes)
)

# Peek a few rows (put key columns first, then everything else)
low_blunt_high_bayes %>%
  relocate(blunt_score, Combined.Probability, diff) %>%
  slice_head(n = 50)

high_blunt_low_bayes %>%
  relocate(blunt_score, Combined.Probability, diff) %>%
  slice_head(n = 50)

```

```{r}
original <- read_csv("../Data/all_voyages.csv")
data <- read_csv("../Data/data_enriched.csv")

nrow(original)
nrow(data)
```

```{r}
sum(data$matchType=="EXACT", na.rm = TRUE)
sum(data$has_species==TRUE, na.rm = TRUE)
sum(is.na(data$nearest_obis_m))
sum(is.na(data$dr))
sum(is.na(data$TotalSpecies))
sum(is.na(data$Pct_GenusDNA_inDB_ANY))

```

